\section{Electric charge neutrality}


A pion condensate will have an electric charge.
In the grand canonical ensemble, the QCD Lagrangian will have the term  $\mu_Q \bar q Q q$, where $\bar q Q q$ is the electric charge density, $Q$ the quark charge matrix \autoref{three-flavor charge matrix}, and $\mu_Q = \mu_I + 2 \mu_S$ is the electric charge chemical potential.
In the case of $\mu_S = 0$, the charge density is thus
%
\begin{equation}
    n_Q = - \pdv{\Eff}{\mu_Q} = n_I.
\end{equation}
%
A realistic astrophysical object will not have a macroscopic electric charge.
We will therefore model pion stars with the additional constraint of charge neutrality, by including charged leptons in the form of muons or electrons.
These leptons are free fermions, with an electric charge of $- e$.
We may therefore use the results from \autoref{section: cold fermi star} and \autoref{section: fermions}.
The electric charge density of the leptons is given by minus particle number $n_\ell$, which we found in \autoref{Fermi gas particle density},
%
\begin{equation}
    n_{\ell} = \frac{8}{3} 
    \frac{u_{\ell, 0}}{m_\ell} x_f^3,
\end{equation}
%
where $x_f = \sqrt{ {\mu_\ell^2}/{m_\ell^2} - 1}$, is the dimensionless fermi momentum, $m_\ell$ the lepton mass, and $\mu_\ell$ the lepton chemical potential.
This formula is valid for $\mu_ell \geq m_\ell$.
We have introduced the characteristic energy density of leptons,
%
\begin{equation}
    u_{\ell, 0} = \frac{m^4_\ell}{3 \pi^2}
\end{equation}
%
The criterion for charge neutrality is thus
%
\begin{equation}
    \label{criterion charge neutrality}
    n_I = n_\ell.
\end{equation}
%
With this, we can determine the lepton chemical potential given a isospoin chemical potential, $\mu_\ell = \mu_\ell(\mu_I)$.
The leading order result for the pion condensate is given in \autoref{pressure leading order chpt}.
Inserting these results into \autoref{criterion charge neutrality}, we get
%
\begin{equation}
    A \left(\frac{\mu_\ell^2 }{m_\ell^2} - 1 \right)^{3/2}
    = \frac{\mu_I}{m_\pi}\left( 1 - \frac{m_\pi^4}{\mu_I^4}  \right),
\end{equation}
%
The right and left side vanish at $(\mu_I, \mu_\ell) = (m_\pi, m_\ell)$, which we have seen earlier is the point where the pressure and energy density of both the Fermi gas and the pion condensate vanish.
This shows that our pion star will have a well-defined radius.
We have introduced the dimensionless constant
%
\begin{equation}
    A = \frac{3}{8} \frac{m_\pi} {m_\ell} \frac{u_{0, \ell}}{u_0}
    = \frac{1}{8 \pi^2} \frac{m_\ell^3}{m_\pi f_\pi^2}.
\end{equation}
%
Setting the lepton mass to the electron mass or muon mass gives, respectively, $A = 1.3599 \times10^{- 7}$ and $A = 5.7092 \times 10^{3}$.
The explicit form of the functional relation is
%
\begin{equation}
    \frac{\mu_\ell}{m_\ell}
    =
    \sqrt{
        1 + A^{-2/3}
        \left[
            \frac{\mu_I}{m_\pi}\left( 1 - \frac{m_\pi^4}{\mu_I^4}  \right)
        \right]^{2/3}
    }.
\end{equation}
%

The total pressure and energy density will now be the sum of the contributions from the pion condensate and the leptons.
The lepton contribution to these, which we found in \autoref{Fermi gas energy density} and \autoref{Fermi gas pressure}, is
%
\begin{align}
    u_\ell 
    &= u_{\ell,0} 
    \left[(2x_f^3 + x_f) \sqrt{1 + x_f^2} - \arcsinh\left(x_f\right)\right], \\
    p_\ell
    &=\frac{1}{3} u_{\ell,0}
    \left[(2x_f^3 - 3x_f) \sqrt{1 + x_f^2} + 3\arcsinh\left(x_f\right)\right].
\end{align}
%
The contribution from the pion condensate is, as before we found in \autoref{pressure leading order chpt} and \autoref{energy density leading order chpt},
%
\begin{align}
    u_\pi &= \frac{1}{2} u_0 \left( \frac{\mu_I^2}{m_\pi^2} - \frac{m_\pi^2}{\mu_I^2}\right)^2 \\
p_\pi &= \frac{1}{2} \left( 2 + \frac{\mu_I^2}{m_\pi^2} - 3 \frac{m_\pi^2}{\mu_I^2}  \right)
\end{align}
%
This leads to a total pressure and energy
%
\begin{equation}
    p = p_\pi + p_\ell, \quad u = u_\pi + u_\ell.
\end{equation}
%
As \autoref{cirterion charge neutrality} gives $\mu_\ell$ as a function of $\mu_I$, these are both parametrized by the isospin chemical potential, and as before we can thus extract the equation of state $u = u(p)$.


